\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-20 paper.tex
save

Complex computer simulators are increasingly used across fields of science to
describe generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely provide a way to evaluate the likehood
function for a given observation. In this note, we develop a likelihood-free
inference algorithm for fitting a forward non-differentiable generative model to
observed data. We adapt the adversarial training procedure of generative
adversarial networks by replacing the implicit generative network with a
domain-based scientific simulator, and solve the resulting non-differentiable
minimax problem with variational optimization. Effectively, the adversarial
variational optimization procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the respective
hierarchical model matches the observed data.
\glnote{Mention experimental results.}
\glnote{Add 'so what?' conclusion.}

@@

In many fields of science such as particle physics, climatology or population
genetics, computer simulators are used to describe complex processes that tie
parameters of an underlying theory to high dimensional observations. In most
cases, these implicit generative models~\citep{2016arXiv161003483M} are
specified as imperative implementations of forward stochastic processes that
generate data. Because it is usually computationally intractable, simulators
often do not provide a way to directly evaluate the likelihood function for a
given observation, thereby making inference difficult.

@@

\subsection{Generative adversarial networks}

@@

As thoroughly explored in \citep{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\citep{2017arXiv170107875A} reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W(p_r, p_\theta)$ by
replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \phi) : \mathbb{R}^d \to \mathbb{R}$. Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\phi$ and $\theta$ in order to respectively minimize
\begin{align}
    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p_\theta} [d(\tilde{\mathbf{x}};\phi)] - \mathbb{E}_{\mathbf{x} \sim p_r} [d(\mathbf{x};\phi)]  \nonumber \\
                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p_{\hat{\mathbf{x}}}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\phi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p_\theta} [d(\tilde{\mathbf{x}};\phi)]
\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} + (1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim p_r$ and $\tilde{\mathbf{x}} \sim p_\theta$.
\glnote{Mention that we use WGAN-GP in this note, but that the VO procedure can be coupled other vairants of GANs.}

@@

The alternating stochastic gradient descent on ${\cal L}_d$ and ${\cal L}_g$ in
GANs inherently assumes that the generator $g$ is a differentiable function. In
the setting where we are not interested in learning the implicit model itself but are
rather interested in the inference of parameters of a fixed
non-differentiable simulator (as outlined in Section~\ref{sec:problem}),
gradients $\nabla_\theta g$ either do not exist or cannot be accessed. As a
result, gradients $\nabla_\theta {\cal L}_g$ cannot be constructed and the
optimization procedure cannot be carried out.

@@

\glnote{add algorithm}

>>>>

Complex computer simulators are increasingly used across fields of science to
describe generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely provide a way to directly evaluate the likehood
function for a given observation. In this note, we develop a likelihood-free
inference algorithm for fitting a forward non-differentiable generative model to
observed data. We adapt the adversarial training procedure of generative
adversarial networks by replacing the implicit generative network with a
domain-based scientific simulator, and solve the resulting non-differentiable
minimax problem with variational optimization. Effectively, the adversarial
variational optimization procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the respective
hierarchical model matches the observed data.
\glnote{Mention experimental results.}
\glnote{Add 'so what?' conclusion.}

@@

In many fields of science such as particle physics, climatology or population
genetics, computer simulators are used to describe complex processes that tie
parameters of an underlying theory to high dimensional observations. In most
cases, these implicit generative models~\citep{2016arXiv161003483M} are
specified as imperative implementations of forward stochastic processes that
generate data. Because it is usually computationally intractable, most simulators
do not provide a way to directly evaluate the likelihood function for a
given observation, thereby making inference difficult.

@@

\subsection{Generative adversarial networks}
\label{sec:gans}

@@

As thoroughly explored in \citep{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\citep{2017arXiv170107875A} reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W(p_r, p_\theta)$ by
replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \phi) : \mathbb{R}^d \to \mathbb{R}$. Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\phi$ and $\theta$ in order to respectively minimize
\begin{align}
    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p_\theta} [d(\tilde{\mathbf{x}};\phi)] - \mathbb{E}_{\mathbf{x} \sim p_r} [d(\mathbf{x};\phi)]  \nonumber \\
                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p_{\hat{\mathbf{x}}}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\phi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p_\theta} [d(\tilde{\mathbf{x}};\phi)]
\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} + (1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim p_r$ and $\tilde{\mathbf{x}} \sim p_\theta$.

@@

\begin{figure*}
    \begin{minipage}{\linewidth}
    \begin{algorithm}[H]
    \caption{Adversarial variational optimization.}

    \begin{flushleft}
        {\it Inputs:} observed data $\{ \mathbf{x}_i \}_{i=1}^N$, simulator $g$;\\
        {\it Outputs:} proposal distribution $q_\psi(\theta)$;\\
        {\it Hyper-parameters:} Number $T$ of training iterations,
                                Number $K$ of gradient steps to update $d$,
                                Size $M$ of the minibatches.
    \end{flushleft}

    \label{alg:avo}
    \begin{algorithmic}[1]
        \State{\glnote{todo!}}
        \For{$t=1$ to $T$}
            % \For{$k=1$ to $K$} \Comment{Update $r$}
            %     \State{Sample minibatch $\{x_m, z_m, s_m = f(x_m;\theta_f) \}_{m=1}^M$} of size $M$;
            %     \State{With $\theta_f$ fixed, update $r$ by ascending its stochastic gradient $\nabla_{\theta_r} E(\theta_f, \theta_r) :=$
            %     $$\nabla_{\theta_r} \sum_{m=1}^M \log p_{\theta_r}(z_m|s_m)  ;$$}
            % \EndFor
            % \State{Sample minibatch $\{x_m, y_m, z_m, s_m = f(x_m;\theta_f)  \}_{m=1}^M$} of size $M$; \Comment{Update $f$}
            % \State{With $\theta_r$ fixed, update $f$ by descending its stochastic gradient $\nabla_{\theta_f} E(\theta_f, \theta_r) :=$
            % $$\nabla_{\theta_f}  \sum_{m=1}^M \left[ -\log p_{\theta_f}(y_m|x_m)  +\log p_{\theta_r}(z_m|s_m)  \right],$$
            % \indent where $p_{\theta_f}(y_m|x_m)$ denotes $\mathbf{1}(y_m=0)(1-s_m) + \mathbf{1}(y_m=1)s_m$;}
        \EndFor
    \end{algorithmic}
    \end{algorithm}
    \end{minipage}
\end{figure*}

The alternating stochastic gradient descent on ${\cal L}_d$ and ${\cal L}_g$ in
GANs (Section~\ref{sec:gans}) inherently assumes that the generator $g$ is a differentiable function. In
the setting where we are not interested in learning the implicit model itself but are
rather interested in the inference of parameters of a fixed
non-differentiable simulator (Section~\ref{sec:problem}),
gradients $\nabla_\theta g$ either do not exist or cannot be accessed. As a
result, gradients $\nabla_\theta {\cal L}_g$ cannot be constructed and the
optimization procedure cannot be carried out.

@@

For completeness, Algorithm~\ref{alg:avo} finally outlines the adversarial variational
optimization procedure when coupled with WGAN-GP.
